---
title: "Palmer Penguins Explorer"
author: "Dhruva Arcot"
format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 2
    code-fold: true
    fig-width: 9
    fig-height: 6
execute:
  warning: false
  message: false
---

```{r}
#| label: setup
#| echo: false
library(tidyverse)
library(palmerpenguins)
```

## Overview

The `palmerpenguins` dataset contains measurements for **`r nrow(drop_na(penguins))`** penguins** across three species — *Adelie*, *Chinstrap*, and *Gentoo* — collected from three islands in the Palmer Archipelago, Antarctica.

```{r}
#| label: summary-table
#| echo: false
penguins %>%
  drop_na() %>%
  group_by(species, island) %>%
  summarise(
    Count          = n(),
    `Avg Flipper (mm)` = round(mean(flipper_length_mm), 1),
    `Avg Mass (g)`     = round(mean(body_mass_g), 0),
    .groups = "drop"
  ) %>%
  knitr::kable(caption = "Summary statistics by species and island")
```

## Flipper Length vs. Body Mass

Larger flippers are a strong predictor of body mass — and the three species occupy almost entirely separate morphological spaces.

```{r}
#| label: scatter-plot
#| echo: false

penguins_clean <- penguins %>% drop_na()

species_medians <- penguins_clean %>%
  group_by(species) %>%
  summarise(
    flipper_length_mm = median(flipper_length_mm),
    body_mass_g       = median(body_mass_g)
  )

species_colors <- c(
  "Adelie"    = "#E8855A",
  "Chinstrap" = "#8AB4C9",
  "Gentoo"    = "#2D6A73"
)

ggplot(penguins_clean, aes(x = flipper_length_mm, y = body_mass_g, color = species)) +
  stat_ellipse(
    aes(fill = species),
    geom      = "polygon",
    alpha     = 0.08,
    linewidth = 0.4,
    linetype  = "dashed"
  ) +
  geom_point(color = "white", size = 3.2, alpha = 0.9) +
  geom_point(size = 2.4, alpha = 0.85) +
  geom_point(data = species_medians, shape = 3, size = 6, stroke = 1.5) +
  geom_text(
    data      = species_medians,
    aes(label = species),
    nudge_y   = 160,
    fontface  = "bold",
    size      = 3.5
  ) +
  scale_color_manual(values = species_colors, guide = "none") +
  scale_fill_manual(values  = species_colors, guide = "none") +
  scale_x_continuous(name = "Flipper Length (mm)", breaks = seq(170, 240, by = 10)) +
  scale_y_continuous(name = "Body Mass (g)", labels = scales::comma_format()) +
  labs(
    title    = "Bigger Flippers, Bigger Birds",
    subtitle = "Flipper length vs. body mass · Crosshairs mark species medians · Ellipses show 95% confidence region",
    caption  = "Data: Gorman et al. (2014) via {palmerpenguins}"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.background    = element_rect(fill = "#EEF3F7", color = NA),
    panel.background   = element_rect(fill = "#EEF3F7", color = NA),
    panel.grid.major.x = element_blank(),
    panel.grid.minor   = element_blank(),
    panel.grid.major.y = element_line(color = "#C5D2DC", linewidth = 0.4),
    axis.text          = element_text(color = "#4A6070", size = 10),
    axis.title         = element_text(color = "#2D4A5A", size = 10.5, face = "bold"),
    plot.title         = element_text(color = "#1A2E38", size = 18, face = "bold",
                                      margin = margin(b = 4)),
    plot.subtitle      = element_text(color = "#4A6070", size = 10,
                                      lineheight = 1.4, margin = margin(b = 16)),
    plot.caption       = element_text(color = "#8AAABB", size = 8, hjust = 0,
                                      margin = margin(t = 12)),
    plot.margin        = margin(20, 24, 16, 20)
  )
```

## Bill Dimensions by Species

Bill shape differs sharply between species — Chinstrap penguins in particular have notably long, narrow bills relative to their size.

```{r}
#| label: bill-plot
#| echo: false

ggplot(penguins_clean, aes(x = bill_length_mm, y = bill_depth_mm, color = species)) +
  stat_ellipse(aes(fill = species), geom = "polygon", alpha = 0.08,
               linewidth = 0.4, linetype = "dashed") +
  geom_point(color = "white", size = 3.2, alpha = 0.9) +
  geom_point(size = 2.4, alpha = 0.85) +
  scale_color_manual(values = species_colors, name = "Species") +
  scale_fill_manual(values  = species_colors, guide = "none") +
  scale_x_continuous(name = "Bill Length (mm)") +
  scale_y_continuous(name = "Bill Depth (mm)") +
  labs(
    title    = "Bill Shape Separates Species Cleanly",
    subtitle = "Gentoo bills are long and shallow; Adelie bills are short and deep",
    caption  = "Data: Gorman et al. (2014) via {palmerpenguins}"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.background    = element_rect(fill = "#EEF3F7", color = NA),
    panel.background   = element_rect(fill = "#EEF3F7", color = NA),
    panel.grid.major.x = element_blank(),
    panel.grid.minor   = element_blank(),
    panel.grid.major.y = element_line(color = "#C5D2DC", linewidth = 0.4),
    axis.text          = element_text(color = "#4A6070", size = 10),
    axis.title         = element_text(color = "#2D4A5A", size = 10.5, face = "bold"),
    plot.title         = element_text(color = "#1A2E38", size = 16, face = "bold",
                                      margin = margin(b = 4)),
    plot.subtitle      = element_text(color = "#4A6070", size = 10,
                                      lineheight = 1.4, margin = margin(b = 16)),
    plot.caption       = element_text(color = "#8AAABB", size = 8, hjust = 0,
                                      margin = margin(t = 12)),
    legend.background  = element_rect(fill = "#EEF3F7", color = NA),
    plot.margin        = margin(20, 24, 16, 20)
  )
```

## Data

The full cleaned dataset (`r nrow(penguins_clean)` observations):

```{r}
#| label: data-table
#| echo: false
penguins_clean %>%
  select(species, island, bill_length_mm, bill_depth_mm,
         flipper_length_mm, body_mass_g, sex, year) %>%
  head(20) %>%
  knitr::kable(
    caption = "First 20 observations (cleaned dataset)",
    col.names = c("Species", "Island", "Bill Length (mm)", "Bill Depth (mm)",
                  "Flipper (mm)", "Mass (g)", "Sex", "Year")
  )
```
